---
title: "Antibiotics Under Our Feet"
format: 
  html:
    theme: 
      light: flatly
      dark: darkly
---

## Midterm format

* Students will be provided with a GitHub Classroom link to fork some starter code and Word document demonstrating what the solution should look like.
* The objective is to replicate the graphical analysis given in the provided Word document.
* Students will be graded on the following aspects of their work:
  * R Code - 40% (renders without errors, completion of all sections, general neatness and documentations, correct files are pushed to GitHub, use of vectorization)
  * Figures - 30% (proper variable transformation, use of color, general neatness and readability)
  * Word document - 15% (submission of Word document, code is not shown, yaml header is correct)
  * Understanding - 15% (choice of transformation, written figure analysis, and conclusions)
* When code is pushed to GitHub a pull request will automatically be opened. Do *not* close the pull requesst. I will give you feedback through the pull request.

## Antibiotics Under Our Feet

The dataset we will be using comes from a study called Antibiotics Under Our Feet[^1]. It has been [shared on figshare](https://figshare.com/articles/dataset/Antibiotics_under_our_feet_investigating_antimicrobial_resistance_Scotland_/23770230).

[^1]: Gul, Anum; Mackenzie, Dylan; Shahid, Aleena; Tawfeeq, Tawfeeq; Cornwell, Rebecca; Czekster, Clarissa Melo (2023). Antibiotics under our feet: investigating antimicrobial resistance (Scotland). figshare. Dataset. <https://doi.org/10.6084/m9.figshare.23770230.v1>

```{r setup}
#| echo: false
#| include: false
library(dplyr)
library(readxl)
library(stringr)
library(purrr)

library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

library(knitr)


# read in taxonomy for each microbe
tax <- read_excel('AUOF investigating antimicrobial resistance student spreadsheet.xlsx',
                  sheet = 'Taxonomy') %>%
  
  # split out taxonomy levels
  mutate(Lineage = map(Lineage, ~ str_split(.x, '; ')[[1]]),
         Domain  = map_chr(Lineage, ~ .x[1]),
         Kingdom = map_chr(Lineage, ~ .x[2]),
         Phylum  = map_chr(Lineage, ~ .x[3]),
         Class   = map_chr(Lineage, ~ .x[4]),
         Order   = map_chr(Lineage, ~ .x[5]),
         Family  = map_chr(Lineage, ~ .x[6]),
         Genus   = map_chr(Lineage, ~ .x[7]))


# read in study data
dat <- read_excel('AUOF investigating antimicrobial resistance student spreadsheet.xlsx',
                  sheet = 2) %>%
  
  # rename columns
  rename(Kingdom = `Classification - bacteria, fungi?`,
         Disease_long = `What disease does it cause?`,
         Resistance = `Antimicrobial that it was resistant to?`,
         Action_long = `Method of action of antimicrobial drug?`,
         Observed_long = `Microbe species present our soil samples?`) %>%
  
         # fix misspelling of Citrobacter freundii
  mutate(Microbe = ifelse(Microbe == 'Citrobacter fruendii', 'Citrobacter freundii', Microbe),
         
         # categorize antimicrobial action
         Action = case_when(grepl(              'peptidoglycan', Action_long) ~ 'Peptidoglycan inhibition',
                            grepl('penicillin-binding proteins', Action_long) ~ 'PBP',
                            grepl( 'penecilin-binding proteins', Action_long) ~ 'PBP',
                            grepl(      'egrosterol|ergosterol', Action_long) ~ 'Ergosterol inhibition',
                            grepl(       'interacting with DNA', Action_long) ~ 'DNA interactions',
                            grepl(              'glucan|gulcan', Action_long) ~ 'Glucan inhibition',
                            grepl(                  'cell wall', Action_long) ~ 'Cell wall disruption',
                            grepl(           'mucopeptide wall', Action_long) ~ 'Cell wall disruption',
                            grepl(              'cell membrane', Action_long) ~ 'Cell wall disruption',
                            grepl(         'ribosome|ribosomal', Action_long) ~ 'Translation disruption',
                            TRUE                                              ~ 'Other'),
         
         # observed into a logical variable
         Observed = grepl('Yes', Observed_long)) %>%
         
  
  # remove "EXAMPLE" row
  filter(!grepl("EXAMPLE", Microbe)) %>%
  
  # merge taxonomy data into the dat
  select(-Kingdom) %>% # drop this, as it 
  left_join(tax)
```

```{r}
#| echo: false
group_by(dat, Action) %>%
  summarize(`Number of Microbes` = n(),
            Observed = sum(Observed)) %>%
  ungroup() %>%
  kable(caption = "Number of microbes with documented antimicrobial resistance that were targetd by this study, grouped by antimicrobial action, and the number of microbes found in study samples.")
```

```{r}
#| echo: false
ggplot(dat, aes(Phylum)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| echo: false
ggplot(dat, aes(Class)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```